Measuring contrast processing in the visual system using SSVEP

### 2024-05-06 review.qmd
Reference update

The SSVEP can also be used to study changes in visual processing \textit{driven} by different behavioural states or overall arousal. For example, locomotion has been shown to alter neuronal excitability and spatial normalization in mice [@NeilStryker2010, @Ayaz2013] - running mice have higher visual sensitivity and lower surround suppression compared to stationary mice. Although measuring EEG responses from locomoting humans is technically challenging, SSVEP studies (which are able to distinguish broadband noise from input signal effectively) have shown that walking also alters early visual processing although in a manner different to that observed in mice [@Benjamin2018, @CaoPlosBiol2019]

SSVEP measurements are typically used to measure time invariant responses due to sustained attention. However, recent work has shown that moderately high modulation frequencies (ca 10Hz) and short analysis windows (1.5s) can also be used to track the dyanmic allocation of attention across a task [@ChotaJCN2024] or attention to moving targets [@deLissaPlos12020]. The ability to track changes in attention over short time periods is also important if SSVEP is to be used for dynamic readouts - for example in a Brain Computer Interface.

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf]. BEcause SSVEP signals are highly sensitive to both visual field position [@Ales2010, @dirusso2007_HBM_spatiotemporal] and attentional state [@hillyard1996_PNAS, @Hillard1998_NNS, @Verghese2012; @Lauritzen2010],  the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is typically present, this modulation and the associated changes in gain control, will affect the entire pattern of self- and intermodulation terms allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties : For example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

@@

*2: Sweeps and CRFs. Measurements of contrast sensitivity, extrapolating
the sweep to zero response to get t'hold. Infants and adults?*

*3: Measurements of modulation. Figures adapted from other papers:
Attention to space, attention to features, adaptation (?).
masking/surround suppression\\*

*4: Clinical (Porciatti / Tsai/ Marmite/ Amblyopia / PD?)*

*5: Future directions (decoding in frequency domain?, animals? BCI?)*

[]{#anchor-2}History

[]{#anchor-3}1. The basics of SSVEP and contrast sensitivity including a
history of both fields

SSVEP (Steady-State Visual Evoked Potential): A continuous electrical
response evoked in the brain by visual stimuli flickering at a constant
frequency (Regan, 1966).

Contrast Sensitivity: The ability to detect differences in luminance
between an object and its background (Campbell \& Green, 1965).

Regan, D. (1966). Some characteristics of average steady-state and
transient responses evoked by modulated light. Electroencephalography
and clinical neurophysiology, 20(3), 238-248.

Campbell, F. W., \& Green, D. G. (1965). Optical and retinal factors
affecting visual resolution. J Physiology, 181, 576-593.

Check Regan paper for earlier (e.g. EEG refs). Norcia review will be
helpful!

(From Tyler / Levi / Apkarian paper):

6\. Regan, D.: Rapid methods for refracting the eye and assessing the
visual acuity in amblyopia using steady-state visual evoked potentials.
In Desmedt, J.E., editor: Visual Evoked Potentials in Man: New
Developments, Oxford, 1977, Clarendon Press, pp. 418-426.

7\. Fricker, S.J.: Narrow-band filter techniques for the detection and
measurement of evoked responses, Electroencephalogr. Clin. Neurophysiol.
14:411, 1962.

8\. Van der Tweel, L. H., Sem-Jacobsen, C.W., Kamp, A., Van Leeuwen,
W.S., and Verings, F.T.H.: Objective determination of response to
modulated light, Acta Physiol. Pharmacol. Neerl. 7:528, 1958.

9\. Regan, D.: Latencies of evoked potentials to flicker and to pattern
speedily estimated by simultaneous stimulation method,
Electroencephalogr. Clin. Neurophysiol. 40:654, 1976.

10\. Tyler, C.W., Apkarian, P., and Nakayama, K.: Multiple spatial
frequency tuning of electrical responses from the human\....

\[Around here perhaps a section about what gain control is, mentioning
other methods as well including psychophysics, MRI, electrophysiology,
and other EEG markers including ERPs and evoked gamma band oscillations.
Maybe outline the Heeger gain control model and its cousins.\]

\[Sure. But in fact gain control is only a small part of this story -
especially in the early days. They were, I think, more interested in
using SSVEP to measure absolute contrast sensitivity and to get lower
bounds on things like infant visual development. Gain control might be
better as a separate section later.\]

[]{#anchor-4}2. Spatial, temporal frequency and contrast sensitivity
measurements

Basically using SSVEP to measure a cortical output amplitude for any
given input contrast. You can vary parameters like SF, TF, position,
color and of course contrast. Early on people realised that you can
'sweep' the stimulus to get a CRF. You broadly get a line in log
contrast space if you do that (Tyler) - then you can extrapolate that
line down to zero response to estimate the threshold. That doesn't
\>quite\< work but it's pretty close.

You can also use this for 'difficult' populations like babies. One
interesting story was about how SSVEP become a replacement for
preferential looking (which was the other way of looking at infant
visual development). See e.g. Davida Teller. SSVEPs allowed people to
make objective measurements of contrast sensitivity development and
deduce that the visual system was more mature (e.g. more functional) in
infancy than previously expected. Also measures of colour sensitivity.
Tinyeyes is based off those measurements. Other people: Tyler, Norcia,
Gunilla H-P, many of the people at SKERI in the 1980s and 90s. Norcia
86,88,90 - mentioned in Regan's nice autobiography :

In parallel of course, people were using frequency tagging to do single
unit work - the 1F vs 2F simple/complex cell classification scheme was
all about this (Lennie and others).

[]{#anchor-5}3. SSVEP in functional localization

Techniques like fMRI have been combined with SSVEP to achieve more
precise spatial localization (Di Russo et al., 2007). See however the
Ales cruciform paper.

Ales and Norcia:\... (showed that people's intuition about V1 upper /
lower v.f reversing polarity wasn't really correct). This feels a little
outside our scope though\... We have also combined SSVEP with source
imaging techniques to probe responses in different cortical locations
(again, Ales papers, some early stuff from Stan K? Appelbaum, Wade,
Norcia figure/ground\...) and then a host of later work from that lab
and others.

[]{#anchor-6}4. SSVEP and contrast gain control including adaptation,
masking, and attention

\[The story goes: Up to the lte 90s people were primarily interested in
measuring contrast sensitivity - the shape of the response function was
assumed to be basically log-linear - and they fit it with straight lines
to extrapolate back to zero response. But then (once Heeger's 1992 paper
had sunk in - see also stuff like Shapley and Victor 1981 [(Shapley \&
Victor, 1981)](https://www.zotero.org/google-docs/?zAmtgj)), people
started thinking about gain control - Candy and Norcia in about 1999,
Porciatti, probably a load of Tyler papers that I don't even know
about\...\]. Probably look in p

And then people worked out that if you can use SSVEP to measure contrast
responses, you can also use it to measure things that modulate contrast
responses. These include adaptation, masking, suppression, attention
(feature and space), clinical things.

[]{#anchor-7}Adaptation:

Continuous exposure to high-contrast patterns reduces contrast
sensitivity, which can be measured using SSVEP (Ross et al., 1989).
Others? Baker recent gain control paper is perhaps worth mentioning here
as a 'confound' of sorts. Engel 2018 [*(Vergeer et al.,
2018)*](https://www.zotero.org/google-docs/?kdURiS). This paper is
interesting [*(Rideaux et al.,
2023)*](https://www.zotero.org/google-docs/?J2EO3d) and \>sort\< of
SSVEP.

[]{#anchor-8}Masking:

High contrast masks can suppress the visibility of low contrast
patterns, which has implications in SSVEP amplitude (Haynes et al.,
2003).

Attention: Directing attention can enhance contrast sensitivity, as
shown in studies using SSVEP (Müller et al., 2006). Also Tsai
(dynamics), Baker / Wade (several), Winawer? I think JW has a nice
dynamic model of normalization with some MEG data. Busse et al cat/human
comparison. Candy and Norcia 2001 JNS [(Candy et al.,
2001)](https://www.zotero.org/google-docs/?7hhs5Z)

Ross, J., Speed, H. D., \& Morgan, M. J. (1989). The effects of
adaptation and masking on incremental thresholds for contrast. Vision
research, 29(2), 205-215.

Haynes, J. D., Roth, G., Stadler, M., \& Heinze, H. J. (2003).
Neuromagnetic correlates of perceived contrast in primary visual cortex.
Journal of Neurophysiology, 89(6), 2655-2666.

Müller, M. M., Picton, T. W., Valdes-Sosa, P., Riera, J.,
Teder-Sälejärvi, W. A., \& Hillyard, S. A. (2006). Effects of spatial
selective attention on the steady-state visual evoked potential in the
20--28 Hz range. Cognitive Brain Research, 24(1), 1-13.

5\. Clinical implications

[]{#anchor-9}Clinical applications (this is a whole section)

Not sure exactly how SSVEP used in clinic. mfVEP?

, e.g., in monitoring visual impairments, tracking neuronal diseases, or
neurofeedback (Norcia et al., 2015).

Tsai epilepsy [(Tsai et al.,
2011)](https://www.zotero.org/google-docs/?AdhH3f). Other photosensitive
epilepsy: Porciatti [(Porciatti et al.,
2000)](https://www.zotero.org/google-docs/?UTBQbV)

Citation:

Migraine (Regan et al). Autism?

Norcia, A. M., Appelbaum, L. G., Ales, J. M., Cottereau, B. R., \&
Rossion, B. (2015). The steady-state visual evoked potential in vision
research: A review. Journal of vision, 15(6), 4-4.

Animal work: Flies (Eliott, West, Himmelberg, Ales, Norcia): Mice/rats
(probably many - can't think off the top of my head - we were working
with a mouse EEG person at UCSF about 12 years ago\...), Monkeys
(Kiorpes?),

[]{#anchor-10}6. Future directions

Use as a readout of modulations. TMS? FUS? Marmite B12 / Fluoxetine /
amblyopia in Rats,

GABA Huang

There is a lot of SSVEP interest these days because of BCIs. I think
it's pretty weak but there is \>so much\< of it that it might be worth
mentioning\...

Advanced signal processing techniques and machine learning can be
integrated to improve SSVEP-based systems (Zhu et al., 2010).

Exploring new clinical and diagnostic applications, understanding
neurological diseases, and developing novel therapeutic interventions.

Citation:

Zhu, D., Bieger, J., Molina, G. G., \& Aarts, R. M. (2010). A survey of
stimulation methods used in SSVEP-based BCIs. Computational intelligence
and neuroscience, 2010.

@@

<!-- old code segments that are no longer used -->

```{python}
#| label: fig-plotspectra
#| fig-cap: "Temporal frequency representations of combining two inputs at frequencies F1=5Hz and F2=7Hz. A pure sine-wave input would generate responses only at F1 and F2. However, the rectification that accompanies early visual processing converts this contrast reversal into a more complex frequency-domain representation with power predominantly at the second harmonics 2F1 and 2F2."
#| fig-cap-location: bottom

#| label: plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain"
#| fig-cap-location: bottom

import matplotlib.pyplot as plt
import numpy as np
import scipy as scp

inputFrequencies_Hz=np.array([5,7]) # We simulate a system with two input frequencies (perhaps contrast-reversing gratings?)

duration_S=1 # All variables in this code have an _[units] where possible

digitisationRate_Hz=1000

inputModulations = np.zeros((len(support_S), len(inputFrequencies_Hz)))
for i, freq in enumerate(inputFrequencies_Hz):
    inputModulations[:, i] = np.sin(2 * np.pi * freq * support_S)

freqCutoff_Hz=30
combinedSignal=np.sum(inputModulations,axis=1)
combinedRectifiedSignal=np.sum(np.abs(inputModulations), axis=1)
FT_combinedSignal=np.fft.fft(combinedSignal)/len(combinedSignal)
FT_combinedRectifiedSignal=np.fft.fft(combinedRectifiedSignal)/len(combinedSignal)
fig,axs=plt.subplots(1,4,figsize=(8,3))
axs[0].plot(support_S,combinedSignal)
axs[0].set_xlabel('Time (s)')
axs[0].set_ylabel('Amplitude (a.u.)')
axs[0].set_title('Input signals:\nlinear combination')

# Create the frequency support for the FFT plot
frequencySupport = np.arange(1,freqCutoff_Hz*duration_S+1) # We only want to plot the frequencies in the FFT up to this point

# Plot the absolute values of the FFT in the second axis as a bar plot
axs[1].bar(frequencySupport, np.abs(FT_combinedSignal[frequencySupport]))
axs[1].set_xlabel('Frequency (Hz)')
axs[1].set_ylabel('Magnitude (a.u.)')
axs[1].set_title('FFT\n linear combination')
# Plot the absolute values of the FFT in the second axis as a bar plot

axs[2].plot(support_S,combinedRectifiedSignal)
axs[2].set_xlabel('Time (s)')
axs[2].set_ylabel('Amplitude (a.u.)')
axs[2].set_title('Sum rectified inputs')

axs[3].bar(frequencySupport, np.abs(FT_combinedRectifiedSignal[frequencySupport]))
axs[3].set_xlabel('Frequency (Hz)')
axs[3].set_ylabel('Magnitude (a.u.)')
axs[3].set_title('FFT of\n sum rectified inputs')

plt.tight_layout()
plt.show()

```{python}
#| label: fig-plotsinewaves
#| fig-cap: "a) A typical SSVEP setup: a screen, a grating, a flicker rate: b) The responses from a single electrode (time domain) c) Frequency domain."
#| fig-cap-location: bottom

import matplotlib.pyplot as plt
import numpy as np
import scipy as scp

inputFrequencies_Hz=np.array([5,7]) # We simulate a system with two input frequencies (perhaps contrast-reversing gratings?)

duration_S=1 # All variables in this code have an _[units] where possible

digitisationRate_Hz=1000

inputModulations = np.zeros((len(support_S), len(inputFrequencies_Hz)))

for i, freq in enumerate(inputFrequencies_Hz):
    inputModulations[:, i] = np.sin(2 * np.pi * freq * support_S)

# Create a figure with two subplots
fig, axs = plt.subplots(len(inputFrequencies_Hz), 1, figsize=(8, 6))

# Plot each input modulation in a separate subplot
for i, freq in enumerate(inputFrequencies_Hz):
    axs[i].plot(support_S, inputModulations[:, i])
    axs[i].set_xlabel('Time (s)')
    axs[i].set_ylabel('Amplitude')
    axs[i].set_title('Frequency: {} Hz'.format(freq))

# Adjust the spacing between subplots
plt.tight_layout()
plt.show()

```{python}
#| label: fig-contrastresponsefunction
#| fig-cap: "Non-linear contrast response function resulting from the population average of two full-wave rectified signals passing through a non-linear transducer. Note the 'supersaturation' at high contrast levels."
#| fig-cap-location: bottom

import matplotlib.pyplot as plt
from matplotlib import colormaps
import numpy as np
import scipy as scp

  axs[0].plot(support_S, inputModulation)

   # Now compute the FT on the signal after it has passed through a single non-linearity. Do this 
   # using the transducer functiona above.
  transducedModulation=simpleTransducerFunction(inputModulation, c50,RMax)
  FT_transducedModulation=np.fft.fft(transducedModulation)/ len(transducedModulation)
  transducedAmplitude = np.abs(FT_transducedModulation[inputFrequencies_Hz*2])
  transducedSignal[contrastIndex]=transducedAmplitude[0]

axs[0].set_title('Sine-wave responses at\n different contrast levels', fontsize=10)

axs[1].plot(contrastRange,initialSignal)
axs[1].set_xlabel('Contrast')
axs[1].set_ylabel('Amplitude (a.u.)')
axs[1].set_title('Linear computation stage', fontsize=10)
axs[1].grid()

axs[2].plot(contrastRange,transducedSignal)
axs[2].set_xlabel('Contrast')
axs[2].set_ylabel('Amplitude (a.u.)')
axs[2].set_title('After nonlinear transducer', fontsize=10)

axs[2].grid()
plt.tight_layout()
plt.show()

```{python}
#| label: fig-intermodulation
#| fig-cap: "Non-linear combination of two independent contrast modulations at F1 and F2 results in 'intermodulation' terms that appear at low-order sums and differences of those frequencies."
#| fig-cap-location: bottom

import matplotlib.pyplot as plt
from matplotlib import colormaps
import numpy as np
import scipy as scp

def simpleTransducerFunction(inputContrast, maskContrast, c50, RMax):
 # Here we model a simple hyperbolic ratio function. Also called a "Naka-Rushton" function.
 # RMax defines the reponse gain of the transducer (how big the biggest output is). 
 # c50 defines the conrast sensitivity. Smaller values are more sensitive.
 # The exponent can also vary - here we fix it to 2
  expnt=2
  inputContrast=np.abs(inputContrast)
  maskContrast=np.abs(maskContrast)

# Define some constants
inputFrequency_Hz=np.array([5,7]) # We simulate a system with two inputs at a single contrast
inputContrasts=np.array([.2 ,.2])
duration_S=1 # All variables in this code have an _[units] where possible
digitisationRate_Hz=1000
freqCutoff_Hz=30
c50=.10
RMax=10

# Make a figure with three subplots
fig, axs = plt.subplots(1,3, figsize=(8, 3))

# Compute waveforms for the 'probe' and 'mask' inputs at the frequencies defined in inputFrequency_Hz
probeModulation = np.sin(2 * np.pi * inputFrequency_Hz[0] * support_S) * inputContrasts[0]
maskModulation =  np.sin(2 * np.pi * inputFrequency_Hz[1] * support_S) * inputContrasts[1]

axs[0].plot(support_S, probeModulation,'k')
axs[0].plot(support_S, maskModulation,'r')
axs[0].set_title('Input modulations', fontsize=10)
axs[0].grid()

# pass both the probe and the mask through the transducer
combinedSignal=simpleTransducerFunction(probeModulation, maskModulation, c50,RMax)

# Compute the FT of the combined signals
FT_combinedSignal=np.abs(np.fft.fft(combinedSignal)/ len(combinedSignal))
axs[1].plot(support_S, combinedSignal,'k')
axs[1].set_title('Combined signal after transducer', fontsize=10)
axs[1].grid()

axs[2].bar(np.arange(1,19), FT_combinedSignal[1:19])

axs[2].grid()
plt.tight_layout()
plt.show()

>>>>

The SSVEP can also be used to study changes in visual processing \textit{driven} by different behavioural states or overall arousal. For example, locomotion has been shown to alter neuronal excitability and spatial normalization in mice [@Niell2010; @Ayaz2013] - running mice have higher visual sensitivity and lower surround suppression compared to stationary mice. Although measuring EEG responses from locomoting humans is technically challenging, SSVEP studies (which are able to distinguish broadband noise from input signal effectively) have shown that walking also alters early visual processing although in a manner different to that observed in mice [@Benjamin2018; @Cao2019].

SSVEP measurements are typically used to measure time invariant responses due to sustained attention. However, recent work has shown that moderately high modulation frequencies (ca 10Hz) and short analysis windows (1.5s) can also be used to track the dynamic allocation of attention across a task [@Chota2024] or attention to moving targets [@Lissa2020]. The ability to track changes in attention over short time periods is also important if SSVEP is to be used for dynamic readouts - for example in a Brain Computer Interface.

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and frequency resolution of SSVEPs make them an ideal candidate for this approach. Typical studies may involve presenting an array of stimuli at different contrast flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010],  the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is typically present, this modulation and the associated changes in gain control, will affect the entire pattern of self- and intermodulation terms allowing the choice to be decoded by a multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties : For example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

@@

@@
